\myTitle

### 2024-05-22 body-text-of-paper.tex
matt revised some pig pieces re: Toni's comments

We illustrate the use of \gscramble{} in three example cases. The first is
a simple simulation to show that, unlike sampling with replacement, the
approach of \gscramble{} does not produce RISPI.
The next two are examples using empirical data sets.
The first of these involves
simulating a small number ($<\,$70) of markers that are fixed, or nearly
fixed, for alternate alleles between two species of trout: steelhead trout ({\em 
Oncorhynchus mykiss}) and cutthroat trout ({\em O. clarkii}). The goal of the 
simulation study is to determine how accurately individuals of different hybrid
categories between the species can be identified, and whether the use of linked 
markers deflates the estimates of uncertainty to an unacceptable degree.  The 
second example involves many more markers ($\approx$14,000) and 
investigates the accuracy for
identifying individuals admixed between closely related groups of feral
pigs in Missouri.

@@

Genetic samples were collected from 160 invasive wild pigs in southeastern
Missouri that were lethally removed through population control efforts
conducted as a component of the Missouri Feral Hog Elimination Partnership
by US Department of Agriculture – Animal Plant Health Inspection Services –
Wildlife Services, Missouri Department of Conservation, and other cooperative
federal and state agencies.
DNA was extracted from hair or tissue (collected from animals at the time of euthanasia) with
commercially available magnetic bead recovery kits (MagMax DNA, Thermo Fisher Scientific).
Genotyping was performed with GeneSeek’s Genomic Profiler for Porcine HD which provides
62,128 biallelic SNP loci that are mapped across the 18 autosomal chromosomes.
We implemented standard genotype quality control measures; specifically, we pruned loci with
call rates $<95\%$ and minor allele frequencies $<5\%$.
We implemented linkage disequilibrium (LD) pruning using a window size of 50 loci and step size
of 5 loci to remove markers above a linkage threshold of $R^2 > 0.5$.

@@

The simulated hybrid classes produce
the expected range of admixture with $F_1$ and $F_2$ individuals having mean $Q=0.5$, and $F_2$ individuals having
more variation (Figure~\ref{fig:pig-sim-qs}A).
Backcrossed individuals have higher ancestry associated with the backcrossed population with mean $Q$
approximately $0.75$ ($\BC_1$) and $0.87$ ($\BC_2$) (Figure~\ref{fig:pig-sim-qs}A). Interestingly, the distribution of the $Q$ values for $F_1$ hybrids varies more across the different genetic clusters than it does for the other
hybrid categories.
When the estimated $Q$ values for the $F_1$ hybrids are separated by the population combinations,
$F_1$ hybrids for Pop1-Pop2 have the least variation around $q=0.5$ (standard deviation
= 0.008), while Pop1-Pop3 and Pop2-Pop3
have higher variation (standard deviation = 0.011-0.016) (Figure~\ref{fig:pig-sim-qs}B). Also, all $Q$
distributions for $F_1$ hybrids, regardless of population combination, center around
$0.50$ except for the two distributions
of Pop3, which have a mean of $Q=0.49$ (Figure~\ref{fig:pig-sim-qs}B).
The differences among these
distributions of $Q$ values for the simulated data reflect the genetic differences observed
in the empirical data, specifically, that genetic differentiation is highest between Pop1 and Pop2 and that
Pop3 is the most admixed population with ancestry contributions from both Pop1 and Pop2.
The results emphasize difficulties that may arise in estimating admixture fractions between
groups that have experienced historical gene flow.

%%%
\begin{figure}
\newcommand{\pigsimsqscap}{\footnotesize (A) The density of the estimated Q values from ADMIXTURE for
the simulated runs are
grouped across all population-combinations for each hybrid class and colored by the $K$ genetic cluster
(e.g., the Pop1 density plot for the $\BC_1$ plots the Pop1 genetic cluster for both
the Pop1-Pop2 and Pop1-Pop3 combinations). The density plots across the $K$
genetic clusters are mostly similar with the most variation among the genetic clusters evident in
the $F_1$ plot. (B) Plotting the density of the estimated Q values from ADMIXTURE for the $F_1$ hybrids,
separated by the population combinations, reveals the least variance in estimated Q values for
Pop1-Pop2 compared to Pop1-Pop3 and Pop2-Pop3. Note that these population combinations group the
reciprocal population combination in each plot (e.g., "Pop1\_Pop2" has simulations
from both Pop1-Pop2 and Pop2-Pop1).}
\includegraphics[width=0.48\textwidth]{figures/simulation-density-multipanel.pdf}
\caption[\pigsimsqscap]{\pigsimsqscap}
\label{fig:pig-sim-qs}
\end{figure}
%%%

@@

\section*{Acknowledgements}
Research was supported in part by the USDA-APHIS-National Feral Swine Damage
Management Program. We are grateful for the numerous Wildlife Services and Missouri
Department of Conservation employees that contributed to genetic sampling of wild pigs
within southeastern Missouri. We thank Yniv Palti for providing flanking sequences and the linkage map for 57K markers for making a recombination map for the steelhead/cutthroat trout example. Writing of this paper was facilitated by the Mobile High Altitude Venue for Ecological Analysis,
Genetics and Statistics, through which ECA, MGD, and TJS were scientists-in-residence on location near
Cheyenne, WY for 3 days in October 2023. This is contribution number mHAVEAGAS-003.
This work utilized the Alpine high performance computing resource at the University of Colorado Boulder.
Alpine is jointly funded by the University of Colorado Boulder, the University of Colorado Anschutz, Colorado State University,
and the National Science Foundation (award 2201538).

>>>>

We illustrate the use of \gscramble{} in three example cases. The first is
a simple simulation to show that, unlike sampling with replacement, the
approach of \gscramble{} does not produce RISPI.
The next two are examples using empirical data sets.
The first of these involves
simulating a small number ($<\,$70) of markers that are fixed, or nearly
fixed, for alternate alleles between two species of trout: steelhead trout ({\em 
Oncorhynchus mykiss}) and cutthroat trout ({\em O. clarkii}). The goal of the 
simulation study is to determine how accurately individuals of different hybrid
categories between the species can be identified, and whether the use of linked 
markers deflates the estimates of uncertainty to an unacceptable degree.  The 
second example involves many more markers ($\approx$14,000) and 
investigates the accuracy for
identifying admixed individuals of different hybrid classes between 
closely related groups of invasive wild 
pigs in Missouri.

@@

Genetic samples were collected from 160 invasive wild pigs in southeastern
Missouri that were lethally removed through population control efforts
conducted as a component of the Missouri Feral Hog Elimination Partnership
by US Department of Agriculture – Animal Plant Health Inspection Services –
Wildlife Services, Missouri Department of Conservation, and other cooperative
federal and state agencies.
DNA was extracted from hair or tissue (collected from animals at the time of control) with
commercially available magnetic bead recovery kits (MagMax DNA, Thermo Fisher Scientific).
Genotyping was performed with GeneSeek’s Genomic Profiler for Porcine HD which provides
62,128 biallelic SNP loci that are mapped across the 18 autosomal chromosomes.
We implemented standard genotype quality control measures; specifically, we pruned loci with
call rates $<95\%$ and minor allele frequencies $<5\%$.
We implemented linkage disequilibrium (LD) pruning using a window size of 50 loci and step size
of 5 loci to remove markers above a linkage threshold of $R^2 > 0.5$.

@@

The simulated hybrid classes of wild pig populations demonstrate how \gscramble{} can be 
used to accurately produce 
the expected variation in the admixture fractions of different hybrid classes. 
$F_1$ and $F_2$ individuals had the expected range of admixture with 
mean $Q=0.5$, and $F_2$ individuals having
more variation (Figure~\ref{fig:pig-sim-qs}A).
Backcrossed individuals have higher ancestry associated with the backcrossed population with mean $Q$
approximately $0.75$ ($\BC_1$) and $0.87$ ($\BC_2$) (Figure~\ref{fig:pig-sim-qs}A). Interestingly, the distribution of the $Q$ values for $F_1$ hybrids varies more across the different genetic clusters than it does for the other
hybrid categories.
When the estimated $Q$ values for the $F_1$ hybrids are separated by the population combinations,
$F_1$ hybrids for Pop1-Pop2 have the least variation around $q=0.5$ (standard deviation
= 0.008), while Pop1-Pop3 and Pop2-Pop3
have higher variation (standard deviation = 0.011-0.016) (Figure~\ref{fig:pig-sim-qs}B). Also, all $Q$
distributions for $F_1$ hybrids, regardless of population combination, center around
$0.50$ except for the two distributions
of Pop3, which have a mean of $Q=0.49$ (Figure~\ref{fig:pig-sim-qs}B).
The differences among these
distributions of $Q$ values for the simulated data reflect the genetic differences observed
in the empirical data, specifically, that genetic differentiation is highest between Pop1 and Pop2 and that
Pop3 is the most admixed population with ancestry contributions from both Pop1 and Pop2.
The results emphasize difficulties that may arise in estimating admixture fractions between
groups that have experienced historical gene flow.

%%%
\begin{figure}
\newcommand{\pigsimsqscap}{\footnotesize (A) The density of the estimated Q values from ADMIXTURE for
the simulated runs of wild pigs are
grouped across all population-combinations for each hybrid class and colored by the $K$ genetic cluster
(e.g., the Pop1 density plot for the $\BC_1$ plots the Pop1 genetic cluster for both
the Pop1-Pop2 and Pop1-Pop3 combinations). The density plots across the $K$
genetic clusters are mostly similar with the most variation among the genetic clusters evident in
the $F_1$ plot. (B) Plotting the density of the estimated Q values from ADMIXTURE for the $F_1$ hybrids,
separated by the population combinations, reveals the least variance in estimated Q values for
Pop1-Pop2 compared to Pop1-Pop3 and Pop2-Pop3. Note that these population combinations group the
reciprocal population combination in each plot (e.g., "Pop1\_Pop2" has simulations
from both Pop1-Pop2 and Pop2-Pop1).}
\includegraphics[width=0.48\textwidth]{figures/simulation-density-multipanel.pdf}
\caption[\pigsimsqscap]{\pigsimsqscap}
\label{fig:pig-sim-qs}
\end{figure}
%%%

@@

\section*{Acknowledgements}
Research was supported in part by the USDA-APHIS-National Feral Swine Damage
Management Program. We are grateful for the numerous Wildlife Services and Missouri
Department of Conservation employees that contributed to genetic sampling of wild pigs
within southeastern Missouri. We thank Yniv Palti for providing flanking sequences and the linkage map for 57K markers for making a recombination map for the steelhead/cutthroat trout example. Writing of this paper was facilitated by the Mobile High Altitude Venue for Ecological Analysis,
Genetics and Statistics, through which ECA, MGD, and TJS were scientists-in-residence on location near
Cheyenne, WY for 3 days in October 2023. This is contribution number mHAVEAGAS-003.
This work utilized the Alpine high performance computing resource at the University of Colorado Boulder.
Alpine is jointly funded by the University of Colorado Boulder, the University of Colorado Anschutz, Colorado State University,
and the National Science Foundation (award 2201538). 
The findings and conclusions in this publication are those of the author(s) and should not be construed to represent any official USDA or U.S. Government determination or policy. This research was supported in part by the U.S. Department of Agriculture, Animal Plant Inspection Service, Wildlife Services, National Wildlife Research Center.
